---
title: "EDAandPP"
format: html
editor: visual
---

```{r setup}
# Loading Required Packages:
library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
library(scales)
library(patchwork)
library(knitr)
library(kableExtra)
library(GGally)
```

# Data Loading and Raw Structure

```{r load-data}
# Loading the Data:
load("airbnbListings.RData")

# Storing the Loaded Data Object:
raw_data <- Airbnb_Excel_Lab_AAMB_Co

# Chhecking the Class and Dimensions of the Data:
class(raw_data)
dim(raw_data)
```

```{r variable-structure}
# Inspection of the Raw Variables Types
str(raw_data)
```

```{r clean-variable-names}
# Cleaning Variable Names:
airbnb <- raw_data |>
  clean_names()

# Display Cleaned Variable Names:
names(airbnb)
```

```{r basic-overview-table}
# Creation of a Basic Overview Table:
basic_overview <- tibble(
  Measure = c(
    "Rows",
    "Columns",
    "Duplicate Rows",
    "Duplicate IDs",
    "Unique Hosts",
    "Unique Neighbourhoods",
    "Unique Property Types",
    "Unique Room Types"
  ),
  Value = c(
    nrow(airbnb),
    ncol(airbnb),
    sum(duplicated(airbnb)),
    sum(duplicated(airbnb$id)),
    n_distinct(airbnb$host_id),
    n_distinct(airbnb$neighbourhood_cleansed),
    n_distinct(airbnb$property_type_proper_trim),
    n_distinct(airbnb$room_type_proper_trim)
  )
)

# Outputting the Table in Copy-and-Paste Ready LaTeX:
basic_overview_latex <- basic_overview %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Basic Overview of the Airbnb Listings Dataset",
    align = c("l", "r")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(basic_overview_latex)
```

# Derived & Redundant Variable Removal:

```{r remove-derived-variables}
# Remove Identifier, Label and Derived / Redundant Variables Prior to Missing Data Assessment:
derived_vars <- c(
  # Identifier and label variables not useful for clustering:
  "id",
  "host_id",
  "host_name_trimmed_proper",
  "picture_url",
  
  # Duplicated or less complete location variable:
  "neighbourhood_proper_trim",
  
  # Derived / redundant host count variable:
  "host_total_listings_count",
  
  # Room type is treated as a simplified / derived listing classification:
  "room_type_proper_trim",
  
  # Bedroom count is treated as a derived / overlapping capacity measure:
  "bedrooms",
  
  # Time-window booking variables derived from booking activity:
  "booked_30",
  "booked_60",
  "booked_90",
  "booked_365",
  
  # Revenue variables derived from price and booking activity:
  "revenue_30",
  "revenue_60",
  "revenue_90",
  "revenue_365",
  
  # Shorter availability windows derived from the same availability concept:
  "availability_30",
  "availability_60",
  "availability_90",
  
  # Review count variants derived from the overall review activity concept:
  "number_of_reviews_ltm",
  "number_of_reviews_l30d",
  "reviews_per_month"
)

# Create the Reduced Working Dataset:
airbnb_reduced <- airbnb |>
  select(-any_of(derived_vars))

# Create a Small Audit Table of Removed Variables:
removed_vars_table <- tibble(
  Removed_Variable = derived_vars,
  Reason = c(
    "Unique listing identifier; retained only for auditability in the raw dataset.",
    "Unique host identifier; arbitrary numeric ID not suitable for distance-based modelling.",
    "Host name label; high-cardinality text field not suitable for clustering.",
    "Image URL; not analytically useful for clustering.",
    "Incomplete location field; information retained through neighbourhood_cleansed.",
    "Redundant host count field; host_listings_count retained.",
    "Simplified room classification; overlapping listing type information retained through property_type_proper_trim.",
    "Overlapping capacity measure; listing size retained through accommodates.",
    "Derived booking activity measure.",
    "Derived booking activity measure.",
    "Derived booking activity measure.",
    "Derived booking activity measure.",
    "Derived revenue measure based on price and bookings.",
    "Derived revenue measure based on price and bookings.",
    "Derived revenue measure based on price and bookings.",
    "Derived revenue measure based on price and bookings.",
    "Short-window version of availability.",
    "Short-window version of availability.",
    "Short-window version of availability.",
    "Derived recent review count.",
    "Derived recent review count.",
    "Derived review frequency measure."
  )
)

# Output Removed Variables Table in Copy-and-Paste Ready LaTeX:
removed_vars_latex <- removed_vars_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Variables Removed Prior to Missing Data Assessment",
    align = c("l", "l"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  ) |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "9cm")

cat(removed_vars_latex)
```

# Highly Correlated Variable Removal:

```{r remove-highly-correlated-variables}
# Identify Highly Correlated Numeric Variables:
numeric_data <- airbnb_reduced |>
  select(where(is.numeric))

# Calculate Absolute Correlation Matrix:
cor_matrix <- numeric_data |>
  cor(use = "pairwise.complete.obs") |>
  abs()

# Convert Correlation Matrix to Long Format:
high_corr_pairs <- as.data.frame(as.table(cor_matrix)) |>
  rename(
    Variable_1 = Var1,
    Variable_2 = Var2,
    Correlation = Freq
  ) |>
  filter(
    Variable_1 != Variable_2,
    Correlation >= 0.90
  ) |>
  mutate(
    Pair_ID = pmap_chr(
      list(as.character(Variable_1), as.character(Variable_2)),
      ~paste(sort(c(..1, ..2)), collapse = "_")
    )
  ) |>
  distinct(Pair_ID, .keep_all = TRUE) |>
  select(-Pair_ID) |>
  arrange(desc(Correlation))

# Specify Variables to Remove Based on High Correlation and Interpretability:
high_corr_vars <- c(
  "beds",
  "bathrooms",
  "review_scores_accuracy",
  "review_scores_cleanliness",
  "review_scores_checkin",
  "review_scores_communication",
  "review_scores_location",
  "review_scores_value"
)

# Create the Further Reduced Working Dataset:
airbnb_reduced <- airbnb_reduced |>
  select(-any_of(high_corr_vars))

# Create a Small Audit Table of Removed Highly Correlated Variables:
high_corr_removed_table <- tibble(
  Removed_Variable = high_corr_vars,
  Retained_Information = c(
    "Property capacity retained through accommodates and bedrooms.",
    "Property size retained through accommodates and bedrooms.",
    "Overall review quality retained through review_scores_rating.",
    "Overall review quality retained through review_scores_rating.",
    "Overall review quality retained through review_scores_rating.",
    "Overall review quality retained through review_scores_rating.",
    "Overall review quality retained through review_scores_rating.",
    "Overall review quality retained through review_scores_rating."
  )
)

# Output Highly Correlated Variable Removal Table in Copy-and-Paste Ready LaTeX:
high_corr_removed_latex <- high_corr_removed_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Highly Correlated Variables Removed Prior to Missing Data Assessment",
    align = c("l", "l"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  ) |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "9cm")

cat(high_corr_removed_latex)

```

# Remaining Variables

```{r remaining-variables-table}
# Create Table of Remaining Variables After Redundancy-Based Removal:
remaining_vars_table <- tibble(
  Variable = names(airbnb_reduced),
  Type = sapply(airbnb_reduced, \(x) class(x)[1])
) |>
  mutate(
    Variable_Number = row_number()
  ) |>
  select(Variable_Number, Variable, Type)

# Output Remaining Variables Table in Copy-and-Paste Ready LaTeX:
remaining_vars_latex <- remaining_vars_table |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Variables Retained After Redundancy-Based Removal",
    align = c("r", "l", "l"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  ) |>
  column_spec(1, width = "2.5cm") |>
  column_spec(2, width = "6cm") |>
  column_spec(3, width = "3cm")

cat(remaining_vars_latex)
```


# Missing Data Assessment

```{r missingness-table-reduced}
# Calculate Missingness by Variable in the Reduced Dataset:
missing_summary <- airbnb_reduced |>
  summarise(across(everything(), ~sum(is.na(.x)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "N_Observations_Missing"
  ) |>
  mutate(
    Pct_Observations_Missing = N_Observations_Missing / nrow(airbnb_reduced),
    Pct_Observations_Missing = percent(Pct_Observations_Missing, accuracy = 0.001)
  ) |>
  arrange(desc(N_Observations_Missing))

missing_summary
```

# Key Numeric Variable Distributions

```{r numeric-distributions, fig.cap="Distributions of retained numeric Airbnb listing variables."}
# Plot Distributions of Retained Numeric Variables:
key_numeric <- c(
  "host_listings_count",
  "accommodates",
  "price",
  "availability_365",
  "number_of_reviews",
  "review_scores_rating"
)

# Set Plot Text Sizes:
plot_title_size <- 16
axis_title_size <- 13
axis_text_size <- 10
strip_text_size <- 11

# Keep Only Numeric Variables that Exist in the Reduced Dataset:
existing_key_numeric <- intersect(key_numeric, names(airbnb_reduced))

# Produce Faceted Histograms of the Retained Numeric Variables:
airbnb_reduced |>
  select(all_of(existing_key_numeric)) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value, fill = variable)) +
  geom_histogram(bins = 30, colour = "white", show.legend = FALSE) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    x = NULL,
    y = "Listings"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.title = element_text(size = axis_title_size),
    axis.text = element_text(size = axis_text_size),
    strip.text = element_text(size = strip_text_size, face = "bold")
  )
```

# Transformation & Standardsation of Numeric Variables:

```{r numeric-transform-standardise}
# Specify Numeric Variables to be Used for Modelling:
numeric_model_vars <- c(
  "host_listings_count",
  "accommodates",
  "price",
  "availability_365",
  "number_of_reviews",
  "review_scores_rating"
)

# Specify Strongly Right-Skewed Variables for Log Transformation:
log_transform_vars <- c(
  "host_listings_count",
  "price",
  "number_of_reviews"
)

# Create Transformed and Standardised Numeric Modelling Dataset:
airbnb_numeric_scaled <- airbnb_reduced |>
  select(all_of(numeric_model_vars)) |>
  mutate(
    across(
      all_of(log_transform_vars),
      ~ log1p(.x),
      .names = "log_{.col}"
    )
  ) |>
  select(
    -all_of(log_transform_vars)
  ) |>
  mutate(
    across(
      everything(),
      ~ as.numeric(scale(.x)),
      .names = "scaled_{.col}"
    )
  ) |>
  select(
    starts_with("scaled_")
  )

# Rename Variables for Cleaner Modelling Output:
names(airbnb_numeric_scaled) <- names(airbnb_numeric_scaled) |>
  str_remove("^scaled_") |>
  str_remove("^log_")
```

# Transformed & Standardised Key Numeric Variable Distributions

```{r transformed-standardised-numeric-distributions, fig.cap="Distributions of transformed and standardised numeric Airbnb listing variables."}
# Plot Distributions of Transformed and Standardised Numeric Variables:

# Set Plot Text Sizes:
plot_title_size <- 16
axis_title_size <- 13
axis_text_size <- 10
strip_text_size <- 11

# Rename Variables for Clearer Plot Headings:
airbnb_numeric_scaled_plot <- airbnb_numeric_scaled |>
  rename(
    `log(host_listings_count)` = host_listings_count,
    accommodates = accommodates,
    `log(price)` = price,
    availability_365 = availability_365,
    `log(number_of_reviews)` = number_of_reviews,
    review_scores_rating = review_scores_rating
  )

# Produce Faceted Histograms of the Transformed and Standardised Numeric Variables:
airbnb_numeric_scaled_plot |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value, fill = variable)) +
  geom_histogram(bins = 30, colour = "white", show.legend = FALSE) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    x = "Standardised Value",
    y = "Listings"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.title = element_text(size = axis_title_size),
    axis.text = element_text(size = axis_text_size),
    strip.text = element_text(size = strip_text_size, face = "bold")
  )
```

# Categorical Variable Summaries


```{r categorical-frequency-tables}
# Create Frequency Tables for Retained Categorical Variables:
cat_vars <- c(
  "neighbourhood_cleansed",
  "property_type_proper_trim"
)

# Keep Only Categorical Variables that Exist in the Reduced Dataset:
existing_cat_vars <- intersect(cat_vars, names(airbnb_reduced))

# Create Top Category Frequency Tables:
cat_frequency_tables <- airbnb_reduced |>
  select(all_of(existing_cat_vars)) |>
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Category"
  ) |>
  count(Variable, Category, sort = TRUE) |>
  group_by(Variable) |>
  mutate(
    Pct = percent(n / sum(n), accuracy = 0.1)
  ) |>
  slice_max(n, n = 10, with_ties = FALSE) |>
  ungroup() |>
  rename(
    Frequency = n,
    Percentage = Pct
  )

# Output Categorical Frequency Tables in Copy-and-Paste Ready LaTeX:
cat_frequency_latex <- cat_frequency_tables |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Top Categories for Retained Categorical Variables",
    align = c("l", "l", "r", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  ) |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "6cm") |>
  column_spec(3, width = "3cm") |>
  column_spec(4, width = "3cm")

cat(cat_frequency_latex)
```

```{r categorical-frequency-plots, fig.cap="Top categories for retained categorical Airbnb variables."}
# Visualise Top Categories for Retained Categorical Variables:

# Set Plot Text Sizes:
plot_title_size <- 16
axis_title_size <- 13
axis_text_size <- 10
strip_text_size <- 11

# Create Frequency Plot for Retained Categorical Variables:
cat_frequency_tables |>
  mutate(
    Category = fct_reorder(Category, Frequency)
  ) |>
  ggplot(aes(x = Category, y = Frequency, fill = Variable)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ Variable, scales = "free_y", ncol = 1) +
  labs(
    x = NULL,
    y = "Number of Listings"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.title = element_text(size = axis_title_size),
    axis.text = element_text(size = axis_text_size),
    strip.text = element_text(size = strip_text_size, face = "bold")
  )
```

```{r categorical-consolidation-dummy-encoding}
# Consolidate Retained Categorical Variables Before Dummy Encoding:

# Specify Retained Categorical Variables:
categorical_model_vars <- c(
  "neighbourhood_cleansed",
  "property_type_proper_trim"
)

# Specify the Number of Major Categories to Retain:
n_major_neighbourhoods <- 10
n_major_property_types <- 10

# Identify Major Neighbourhood Categories:
major_neighbourhoods <- airbnb_reduced |>
  count(neighbourhood_cleansed, sort = TRUE) |>
  slice_head(n = n_major_neighbourhoods) |>
  pull(neighbourhood_cleansed)

# Identify Major Property Type Categories:
major_property_types <- airbnb_reduced |>
  count(property_type_proper_trim, sort = TRUE) |>
  slice_head(n = n_major_property_types) |>
  pull(property_type_proper_trim)

# Create Consolidated Categorical Variables:
airbnb_categorical_consolidated <- airbnb_reduced |>
  mutate(
    neighbourhood_grouped = if_else(
      neighbourhood_cleansed %in% major_neighbourhoods,
      neighbourhood_cleansed,
      "Other"
    ),
    property_type_grouped = if_else(
      property_type_proper_trim %in% major_property_types,
      property_type_proper_trim,
      "Other"
    )
  ) |>
  select(
    neighbourhood_grouped,
    property_type_grouped
  )

# Create Audit Table Showing Consolidated Category Frequencies:
consolidated_cat_summary <- airbnb_categorical_consolidated |>
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Category"
  ) |>
  count(Variable, Category, sort = TRUE) |>
  group_by(Variable) |>
  mutate(
    Percentage = percent(n / sum(n), accuracy = 0.1)
  ) |>
  ungroup() |>
  rename(
    Frequency = n
  )

# Output Consolidated Category Summary in Copy-and-Paste Ready LaTeX:
consolidated_cat_summary_latex <- consolidated_cat_summary |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Consolidated Categorical Variables Prior to Dummy Encoding",
    align = c("l", "l", "r", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  ) |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "6cm") |>
  column_spec(3, width = "3cm") |>
  column_spec(4, width = "3cm")

cat(consolidated_cat_summary_latex)

# Dummy Encode the Consolidated Categorical Variables:
airbnb_categorical_dummies <- model.matrix(
  ~ neighbourhood_grouped + property_type_grouped - 1,
  data = airbnb_categorical_consolidated
) |>
  as.data.frame()

# Clean Dummy Variable Names:
names(airbnb_categorical_dummies) <- names(airbnb_categorical_dummies) |>
  str_replace_all("neighbourhood_grouped", "neighbourhood_") |>
  str_replace_all("property_type_grouped", "property_type_") |>
  str_replace_all("[^A-Za-z0-9_]", "_") |>
  str_replace_all("_+", "_") |>
  str_remove("_$") |>
  str_to_lower()

# Combine Standardised Numeric Variables and Dummy-Encoded Categorical Variables:
airbnb_model_data <- bind_cols(
  airbnb_numeric_scaled,
  airbnb_categorical_dummies
)

# Check Final Modelling Dataset Dimensions:
model_data_overview <- tibble(
  Dataset = "Final modelling dataset",
  Rows = nrow(airbnb_model_data),
  Columns = ncol(airbnb_model_data)
)

# Output Final Modelling Dataset Overview in Copy-and-Paste Ready LaTeX:
model_data_overview_latex <- model_data_overview |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Final Modelling Dataset After Categorical Consolidation and Dummy Encoding",
    align = c("l", "r", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(model_data_overview_latex)

```

# Plausibility and Anomaly Checks

```{r anomaly-checks}
# Define Basic Plausibility Flags for the Reduced Dataset:
anomaly_flags <- airbnb_reduced |>
  mutate(
    flag_non_positive_price = !is.na(price) & price <= 0,
    flag_non_positive_accommodates = !is.na(accommodates) & accommodates <= 0,
    flag_negative_host_listings_count = !is.na(host_listings_count) & host_listings_count < 0,
    flag_negative_number_of_reviews = !is.na(number_of_reviews) & number_of_reviews < 0,
    flag_availability_365_outside_range = !is.na(availability_365) & 
      (availability_365 < 0 | availability_365 > 365),
    flag_review_rating_outside_range = !is.na(review_scores_rating) & 
      (review_scores_rating < 0 | review_scores_rating > 5),
    flag_latitude_outside_cpt_range = !is.na(latitude) & 
      (latitude < -35 | latitude > -32),
    flag_longitude_outside_cpt_range = !is.na(longitude) & 
      (longitude < 17 | longitude > 20)
  )

# Summarise Plausibility Flags:
anomaly_summary <- anomaly_flags |>
  summarise(
    across(
      starts_with("flag_"),
      ~sum(.x, na.rm = TRUE)
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Anomaly_Check",
    values_to = "N_Flagged"
  ) |>
  mutate(
    Anomaly_Check = Anomaly_Check |>
      str_remove("^flag_") |>
      str_replace_all("_", " ") |>
      str_to_title()
  ) |>
  arrange(desc(N_Flagged))

# Check that Dummy Variables are Binary:
dummy_binary_summary <- airbnb_categorical_dummies |>
  summarise(
    across(
      everything(),
      ~sum(!(.x %in% c(0, 1)), na.rm = TRUE)
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Dummy_Variable",
    values_to = "N_Non_Binary"
  ) |>
  summarise(
    Anomaly_Check = "Non-Binary Dummy Variable Values",
    N_Flagged = sum(N_Non_Binary)
  )

# Combine Raw Variable and Dummy Variable Checks:
anomaly_summary <- bind_rows(
  anomaly_summary,
  dummy_binary_summary
)

anomaly_summary
```

```{r boxplots, fig.cap="Boxplots used to identify skewness and potential extreme values in retained numeric variables."}
# Boxplots for Retained Numeric Variables:

# Specify Retained Numeric Variables for Outlier Inspection:
outlier_vars <- c(
  "host_listings_count",
  "accommodates",
  "price",
  "availability_365",
  "number_of_reviews",
  "review_scores_rating"
)

# Keep Only Numeric Variables that Exist in the Reduced Dataset:
existing_outlier_vars <- intersect(outlier_vars, names(airbnb_reduced))

# Set Plot Text Sizes:
plot_title_size <- 16
axis_title_size <- 13
axis_text_size <- 10
strip_text_size <- 11

# Produce Faceted Boxplots for Retained Numeric Variables:
airbnb_reduced |>
  select(all_of(existing_outlier_vars)) |>
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Value"
  ) |>
  ggplot(aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(outlier.alpha = 0.25, show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ Variable, scales = "free", ncol = 2) +
  labs(
    x = NULL,
    y = NULL,
    title = "Potential Outliers in Retained Numeric Variables"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.title = element_text(size = axis_title_size),
    axis.text = element_text(size = axis_text_size),
    strip.text = element_text(size = strip_text_size, face = "bold"),
    axis.text.y = element_blank()
  )
```

# Summary of Required Pre-processing

The modelling stage should use the cleaned dataset rather than the raw extract. Identifier and descriptive variables such as listing IDs, picture URLs, host names and raw text labels should be excluded because they do not provide meaningful distance-based structure. Monetary and count variables should be log-transformed where they are strongly right-skewed. Invalid values such as non-positive prices, impossible availability or booking values, and implausible coordinates should either be removed or corrected only if a reliable correction rule exists. Missing values should be handled before clustering because distance-based methods cannot use incomplete rows. Categorical variables such as room type, property type and neighbourhood should be converted to factors, rare categories should be grouped, and dummy variables should be created. Finally, all numeric and dummy variables used for hierarchical clustering, non-hierarchical clustering and SOMs should be standardised so that variables measured on larger scales do not dominate the clustering solution.
