---
title: "EDAandPP"
format: html
editor: visual
---

```{r setup}
# Loading Required Packages:
library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
library(scales)
library(patchwork)
library(knitr)
library(kableExtra)
library(GGally)
```

# Data Loading and Raw Structure

```{r load-data}
# Loading the Data:
load("airbnbListings.RData")

# Storing the Loaded Data Object:
raw_data <- Airbnb_Excel_Lab_AAMB_Co

# Chhecking the Class and Dimensions of the Data:
class(raw_data)
dim(raw_data)
```

```{r variable-structure}
# Inspection of the Raw Variables Types
str(raw_data)
```

```{r clean-variable-names}
# Cleaning Variable Names:
airbnb <- raw_data |>
  clean_names()

# Display Cleaned Variable Names:
names(airbnb)
```

```{r basic-overview-table}
# Creation of a Basic Overview Table:
basic_overview <- tibble(
  Measure = c(
    "Rows",
    "Columns",
    "Duplicate Rows",
    "Duplicate IDs",
    "Unique Hosts",
    "Unique Neighbourhoods",
    "Unique Property Types",
    "Unique Room Types"
  ),
  Value = c(
    nrow(airbnb),
    ncol(airbnb),
    sum(duplicated(airbnb)),
    sum(duplicated(airbnb$id)),
    n_distinct(airbnb$host_id),
    n_distinct(airbnb$neighbourhood_cleansed),
    n_distinct(airbnb$property_type_proper_trim),
    n_distinct(airbnb$room_type_proper_trim)
  )
)

# Outputting the Table in Copy-and-Paste Ready LaTeX:
basic_overview_latex <- basic_overview %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Basic Overview of the Airbnb Listings Dataset",
    align = c("l", "r")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(basic_overview_latex)
```

# Missing Data Assessment

```{r missingness-table}
# Calculate Missingness by Variable:
missing_summary <- airbnb |>
  summarise(across(everything(), ~sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_N") |>
  mutate(
    Missing_Pct = Missing_N / nrow(airbnb),
    Missing_Pct = percent(Missing_Pct, accuracy = 0.1)
  ) |>
  arrange(desc(Missing_N))

# Output Missingness Summary Table in Copy-and-Paste Ready LaTeX:
missing_summary_latex <- missing_summary |>
  filter(Missing_N > 0) |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Variables with Missing Values",
    align = c("l", "r", "r")
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(missing_summary_latex)
```

# Numeric Variable Summaries

```{r numeric-distributions, fig.cap="Distributions of key numeric Airbnb listing variables."}
# PLOT DISTRIBUTIONS OF KEY NUMERIC VARIABLES
key_numeric <- c(
  "price", "accommodates", "bathrooms", "bedrooms", "beds",
  "number_of_reviews", "reviews_per_month", "review_scores_rating",
  "availability_365", "booked_365", "revenue_365"
)

existing_key_numeric <- intersect(key_numeric, names(airbnb))

airbnb %>%
  select(all_of(existing_key_numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  labs(
    x = NULL,
    y = "Listings",
    title = "Distributions of Key Numeric Variables"
  ) +
  theme_minimal()
```

```{r price-transformation-check, fig.cap="Raw and log-transformed price distributions."}
# COMPARE RAW PRICE AND LOG PRICE DISTRIBUTIONS
p1 <- airbnb %>%
  ggplot(aes(x = price)) +
  geom_histogram(bins = 40) +
  labs(title = "Raw Price", x = "Price", y = "Listings") +
  theme_minimal()

p2 <- airbnb %>%
  filter(!is.na(price), price > 0) %>%
  mutate(log_price = log1p(price)) %>%
  ggplot(aes(x = log_price)) +
  geom_histogram(bins = 40) +
  labs(title = "Log-Transformed Price", x = "log(1 + price)", y = "Listings") +
  theme_minimal()

p1 + p2
```

# Categorical Variable Summaries

```{r categorical-frequency-tables}
# CREATE FREQUENCY TABLES FOR MAIN CATEGORICAL VARIABLES
cat_vars <- c(
  "neighbourhood_cleansed",
  "property_type_proper_trim",
  "room_type_proper_trim"
)

existing_cat_vars <- intersect(cat_vars, names(airbnb))

cat_frequency_tables <- airbnb %>%
  select(all_of(existing_cat_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "category") %>%
  count(variable, category, sort = TRUE) %>%
  group_by(variable) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  slice_max(n, n = 10, with_ties = FALSE) %>%
  ungroup()

cat_frequency_tables %>%
  kable(caption = "Top Categories for Main Categorical Variables") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r categorical-frequency-plots, fig.cap="Top categories for main categorical Airbnb variables."}
# VISUALISE TOP CATEGORIES FOR EACH MAIN CATEGORICAL VARIABLE
cat_frequency_tables %>%
  mutate(category = fct_reorder(category, n)) %>%
  ggplot(aes(x = category, y = n)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~variable, scales = "free_y") +
  labs(
    x = NULL,
    y = "Number of listings",
    title = "Top Categories by Frequency"
  ) +
  theme_minimal()
```

# Plausibility and Anomaly Checks

```{r anomaly-checks}
# DEFINE BASIC PLAUSIBILITY FLAGS
anomaly_flags <- airbnb %>%
  mutate(
    flag_non_positive_price = !is.na(price) & price <= 0,
    flag_non_positive_accommodates = !is.na(accommodates) & accommodates <= 0,
    flag_negative_bathrooms = !is.na(bathrooms) & bathrooms < 0,
    flag_negative_bedrooms = !is.na(bedrooms) & bedrooms < 0,
    flag_negative_beds = !is.na(beds) & beds < 0,
    flag_availability_365_outside_range = !is.na(availability_365) & (availability_365 < 0 | availability_365 > 365),
    flag_booked_365_outside_range = !is.na(booked_365) & (booked_365 < 0 | booked_365 > 365),
    flag_review_rating_outside_range = !is.na(review_scores_rating) & (review_scores_rating < 0 | review_scores_rating > 5),
    flag_latitude_outside_cpt_range = !is.na(latitude) & (latitude < -35 | latitude > -32),
    flag_longitude_outside_cpt_range = !is.na(longitude) & (longitude < 17 | longitude > 20)
  )

anomaly_summary <- anomaly_flags %>%
  summarise(across(starts_with("flag_"), ~sum(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "anomaly_check", values_to = "n_flagged") %>%
  arrange(desc(n_flagged))

anomaly_summary %>%
  kable(caption = "Plausibility and Anomaly Checks") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r outlier-table}
# IDENTIFY EXTREME VALUES USING THE 1.5 IQR RULE FOR SELECTED VARIABLES
outlier_vars <- c(
  "price", "accommodates", "bathrooms", "bedrooms", "beds",
  "number_of_reviews", "reviews_per_month", "revenue_365"
)

existing_outlier_vars <- intersect(outlier_vars, names(airbnb))

outlier_summary <- airbnb %>%
  select(all_of(existing_outlier_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  mutate(
    q1 = quantile(value, 0.25, na.rm = TRUE),
    q3 = quantile(value, 0.75, na.rm = TRUE),
    iqr = q3 - q1,
    lower_bound = q1 - 1.5 * iqr,
    upper_bound = q3 + 1.5 * iqr,
    is_outlier = value < lower_bound | value > upper_bound
  ) %>%
  summarise(
    lower_bound = first(lower_bound),
    upper_bound = first(upper_bound),
    outlier_n = sum(is_outlier, na.rm = TRUE),
    outlier_pct = percent(outlier_n / sum(!is.na(value)), accuracy = 0.1),
    .groups = "drop"
  )

outlier_summary %>%
  mutate(across(where(is.numeric), ~round(.x, 2))) %>%
  kable(caption = "Potential Outliers Based on the 1.5 IQR Rule") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r boxplots, fig.cap="Boxplots used to identify skewness and potential extreme values."}
# BOXPLOTS FOR SELECTED NUMERIC VARIABLES
airbnb %>%
  select(all_of(existing_outlier_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(outlier.alpha = 0.25) +
  coord_flip() +
  facet_wrap(~variable, scales = "free", ncol = 2) +
  labs(
    x = NULL,
    y = NULL,
    title = "Potential Outliers in Key Numeric Variables"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_blank())
```

# Location and Ward-Level Exploration

```{r location-plot, fig.cap="Spatial distribution of Airbnb listings across Cape Town."}
# PLOT LISTING LOCATIONS USING LATITUDE AND LONGITUDE
airbnb %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  ggplot(aes(x = longitude, y = latitude)) +
  geom_point(alpha = 0.25, size = 0.6) +
  coord_fixed() +
  labs(
    x = "Longitude",
    y = "Latitude",
    title = "Spatial Distribution of Airbnb Listings"
  ) +
  theme_minimal()
```

```{r neighbourhood-level-averages}
# CALCULATE NEIGHBOURHOOD / WARD LEVEL AVERAGES
ward_level_averages <- airbnb %>%
  group_by(neighbourhood_cleansed) %>%
  summarise(
    listings = n(),
    avg_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE),
    avg_accommodates = mean(accommodates, na.rm = TRUE),
    avg_review_rating = mean(review_scores_rating, na.rm = TRUE),
    avg_availability_365 = mean(availability_365, na.rm = TRUE),
    avg_booked_365 = mean(booked_365, na.rm = TRUE),
    avg_revenue_365 = mean(revenue_365, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(listings))

ward_level_averages %>%
  slice_head(n = 15) %>%
  mutate(across(where(is.numeric), ~round(.x, 2))) %>%
  kable(caption = "Neighbourhood / Ward-Level Averages for the Largest Listing Areas") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r ward-price-plot, fig.cap="Average prices for neighbourhoods / wards with the most Airbnb listings."}
# PLOT AVERAGE PRICE FOR THE LARGEST NEIGHBOURHOODS / WARDS
ward_level_averages %>%
  slice_max(listings, n = 15) %>%
  mutate(neighbourhood_cleansed = fct_reorder(neighbourhood_cleansed, avg_price)) %>%
  ggplot(aes(x = neighbourhood_cleansed, y = avg_price)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Average price",
    title = "Average Price by High-Volume Neighbourhood / Ward"
  ) +
  theme_minimal()
```

# Relationships Between Numeric Variables

```{r correlation-matrix}
# SELECT NUMERIC VARIABLES THAT ARE PLAUSIBLY USEFUL FOR CLUSTERING
cluster_numeric_candidates <- airbnb %>%
  select(
    price, accommodates, bathrooms, bedrooms, beds,
    booked_30, booked_60, booked_90, booked_365,
    revenue_30, revenue_60, revenue_90, revenue_365,
    availability_30, availability_60, availability_90, availability_365,
    number_of_reviews, number_of_reviews_ltm, number_of_reviews_l30d,
    review_scores_rating, review_scores_accuracy, review_scores_cleanliness,
    review_scores_checkin, review_scores_communication,
    review_scores_location, review_scores_value, reviews_per_month,
    host_listings_count, host_total_listings_count
  )

cor_matrix <- cluster_numeric_candidates %>%
  cor(use = "pairwise.complete.obs")

round(cor_matrix, 2) %>%
  kable(caption = "Correlation Matrix for Candidate Numeric Clustering Variables") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 6)
```

```{r correlation-plot, fig.cap="Correlation heatmap for candidate numeric clustering variables."}
# VISUALISE CORRELATIONS AMONG CANDIDATE NUMERIC VARIABLES
cor_df <- as.data.frame(as.table(cor_matrix)) %>%
  rename(var1 = Var1, var2 = Var2, correlation = Freq)

cor_df %>%
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  labs(
    x = NULL,
    y = NULL,
    fill = "Correlation",
    title = "Correlation Heatmap for Candidate Numeric Variables"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6)
  )
```

```{r scatterplot-sample, fig.cap="Pairwise relationships for selected transformed numeric variables."}
# PLOT SELECTED PAIRWISE RELATIONSHIPS USING A SAMPLE
set.seed(123)
pair_data <- airbnb %>%
  transmute(
    log_price = log1p(price),
    accommodates,
    bedrooms,
    availability_365,
    booked_365,
    log_reviews = log1p(number_of_reviews),
    review_scores_rating
  ) %>%
  drop_na() %>%
  slice_sample(n = min(1000, n()))

ggpairs(pair_data)
```

# Optional Dimension Reduction Check

```{r pca-prep}
# PREPARE NUMERIC DATA FOR PCA DIAGNOSTIC ONLY
pca_data <- airbnb %>%
  transmute(
    log_price = log1p(price),
    accommodates,
    bathrooms,
    bedrooms,
    beds,
    availability_365,
    booked_365,
    log_reviews = log1p(number_of_reviews),
    review_scores_rating,
    reviews_per_month = log1p(reviews_per_month),
    host_listings_count = log1p(host_listings_count)
  ) %>%
  drop_na()

pca_fit <- prcomp(pca_data, center = TRUE, scale. = TRUE)

pca_variance <- tibble(
  component = paste0("PC", seq_along(pca_fit$sdev)),
  variance_explained = pca_fit$sdev^2 / sum(pca_fit$sdev^2),
  cumulative_variance = cumsum(variance_explained)
)

pca_variance %>%
  mutate(across(where(is.numeric), ~percent(.x, accuracy = 0.1))) %>%
  kable(caption = "PCA Variance Explained for Numeric Diagnostic Variables") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r pca-scree, fig.cap="Scree plot showing cumulative variance explained by diagnostic principal components."}
# SCREE PLOT FOR PCA DIAGNOSTIC
pca_variance %>%
  ggplot(aes(x = seq_along(variance_explained), y = cumulative_variance)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Number of principal components",
    y = "Cumulative variance explained",
    title = "PCA Scree Plot"
  ) +
  theme_minimal()
```

# Pre-processing Specification for the Main Analysis

Based on the EDA above, the following pre-processing steps should be applied before the clustering and SOM modelling stages.

```{r preprocessing-pipeline}
# CREATE MODELLING DATASET WITH CLEANED AND DERIVED VARIABLES
model_data <- airbnb %>%
  mutate(
    # BASIC CLEANING FLAGS
    valid_price = !is.na(price) & price > 0,
    valid_accommodates = !is.na(accommodates) & accommodates > 0,
    valid_coordinates = !is.na(latitude) & !is.na(longitude),
    
    # LOG TRANSFORM HEAVILY SKEWED MONETARY AND COUNT VARIABLES
    log_price = log1p(price),
    log_revenue_30 = log1p(revenue_30),
    log_revenue_60 = log1p(revenue_60),
    log_revenue_90 = log1p(revenue_90),
    log_revenue_365 = log1p(revenue_365),
    log_reviews = log1p(number_of_reviews),
    log_reviews_ltm = log1p(number_of_reviews_ltm),
    log_reviews_l30d = log1p(number_of_reviews_l30d),
    log_reviews_per_month = log1p(reviews_per_month),
    log_host_listings_count = log1p(host_listings_count),
    
    # CREATE OCCUPANCY / BOOKING INTENSITY FEATURES
    booked_rate_30 = booked_30 / 30,
    booked_rate_60 = booked_60 / 60,
    booked_rate_90 = booked_90 / 90,
    booked_rate_365 = booked_365 / 365,
    availability_rate_365 = availability_365 / 365,
    
    # CONVERT CATEGORICAL VARIABLES TO FACTORS
    neighbourhood_cleansed = as.factor(neighbourhood_cleansed),
    property_type_proper_trim = as.factor(property_type_proper_trim),
    room_type_proper_trim = as.factor(room_type_proper_trim)
  ) %>%
  filter(
    valid_price,
    valid_accommodates,
    valid_coordinates,
    !is.na(neighbourhood_cleansed),
    !is.na(property_type_proper_trim),
    !is.na(room_type_proper_trim)
  )

# COMBINE RARE CATEGORIES BEFORE DUMMY VARIABLE CREATION
model_data <- model_data %>%
  mutate(
    property_type_grouped = fct_lump_min(property_type_proper_trim, min = 50, other_level = "Other"),
    neighbourhood_grouped = fct_lump_min(neighbourhood_cleansed, min = 50, other_level = "Other")
  )
```

```{r final-clustering-matrix}
# SELECT VARIABLES FOR CLUSTERING AND SOM INPUT
# EXCLUDE IDENTIFIERS, RAW TEXT FIELDS, URLS, HOST NAMES AND HIGHLY REDUNDANT RAW VARIABLES
cluster_data <- model_data %>%
  transmute(
    log_price,
    accommodates,
    bathrooms,
    bedrooms,
    beds,
    booked_rate_30,
    booked_rate_90,
    booked_rate_365,
    availability_rate_365,
    log_revenue_365,
    log_reviews,
    log_reviews_per_month,
    review_scores_rating,
    review_scores_cleanliness,
    review_scores_location,
    review_scores_value,
    log_host_listings_count,
    room_type_proper_trim,
    property_type_grouped,
    neighbourhood_grouped
  )

# CREATE DUMMY VARIABLES FOR CATEGORICAL INPUTS
cluster_matrix <- cluster_data %>%
  fastDummies::dummy_cols(
    select_columns = c("room_type_proper_trim", "property_type_grouped", "neighbourhood_grouped"),
    remove_first_dummy = TRUE,
    remove_selected_columns = TRUE
  ) %>%
  drop_na()

# STANDARDISE ALL VARIABLES SO THAT SCALE DOES NOT DRIVE DISTANCE-BASED METHODS
cluster_matrix_scaled <- cluster_matrix %>%
  mutate(across(everything(), scale))

# FINAL MODELLING DIMENSION
final_modelling_overview <- tibble(
  final_rows = nrow(cluster_matrix_scaled),
  final_columns = ncol(cluster_matrix_scaled),
  missing_values_remaining = sum(is.na(cluster_matrix_scaled))
)

final_modelling_overview %>%
  kable(caption = "Final Pre-processed Modelling Matrix Overview") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

# Summary of Required Pre-processing

The modelling stage should use the cleaned dataset rather than the raw extract. Identifier and descriptive variables such as listing IDs, picture URLs, host names and raw text labels should be excluded because they do not provide meaningful distance-based structure. Monetary and count variables should be log-transformed where they are strongly right-skewed. Invalid values such as non-positive prices, impossible availability or booking values, and implausible coordinates should either be removed or corrected only if a reliable correction rule exists. Missing values should be handled before clustering because distance-based methods cannot use incomplete rows. Categorical variables such as room type, property type and neighbourhood should be converted to factors, rare categories should be grouped, and dummy variables should be created. Finally, all numeric and dummy variables used for hierarchical clustering, non-hierarchical clustering and SOMs should be standardised so that variables measured on larger scales do not dominate the clustering solution.
