---
title: "HC"
format: html
editor: visual
---

# Hierarchical Cluster Analysis

This section carries out agglomerative hierarchical clustering on the pre-processed Airbnb modelling dataset. The analysis follows the lecture methodology: first defining a dissimilarity matrix, then comparing linkage rules, assessing the dendrogram fit and candidate values of (K), cutting the tree, and finally profiling the resulting clusters.

```{r setup}
# Loading Required Packages:
library(tidyverse)
library(janitor)
library(cluster)
library(factoextra)
library(scales)
library(knitr)
library(kableExtra)
library(patchwork)

# Set Seed for Reproducibility:
set.seed(2026)
```

## Data Loading and Pre-Processing

The hierarchical clustering analysis uses the same model-ready dataset created during the EDA and pre-processing stage. The raw Airbnb data are loaded, identifiers, labels, derived variables and highly correlated variables are removed, skewed numeric variables are transformed, numeric variables are standardised, and the retained categorical variables are consolidated and dummy encoded.

```{r load-and-preprocess-data}
# Loading the Raw Airbnb Data:
load("airbnbListings.RData")

# Store and Clean the Raw Dataset:
raw_data <- Airbnb_Excel_Lab_AAMB_Co

airbnb <- raw_data |>
  clean_names()

# Remove Identifier, Label, Derived and Redundant Variables:
derived_vars <- c(
  "id",
  "host_id",
  "host_name_trimmed_proper",
  "picture_url",
  "neighbourhood_proper_trim",
  "host_total_listings_count",
  "room_type_proper_trim",
  "bedrooms",
  "booked_30",
  "booked_60",
  "booked_90",
  "booked_365",
  "revenue_30",
  "revenue_60",
  "revenue_90",
  "revenue_365",
  "availability_30",
  "availability_60",
  "availability_90",
  "number_of_reviews_ltm",
  "number_of_reviews_l30d",
  "reviews_per_month"
)

airbnb_reduced <- airbnb |>
  select(-any_of(derived_vars))

# Remove Highly Correlated Numeric Variables:
high_corr_vars <- c(
  "beds",
  "bathrooms",
  "review_scores_accuracy",
  "review_scores_cleanliness",
  "review_scores_checkin",
  "review_scores_communication",
  "review_scores_location",
  "review_scores_value"
)

airbnb_reduced <- airbnb_reduced |>
  select(-any_of(high_corr_vars))

# Retain Raw Variables for Cluster Profiling Later:
airbnb_profile_data <- airbnb_reduced

# Specify Numeric Variables to be Used for Modelling:
numeric_model_vars <- c(
  "host_listings_count",
  "accommodates",
  "price",
  "availability_365",
  "number_of_reviews",
  "review_scores_rating"
)

# Specify Strongly Right-Skewed Variables for Log Transformation:
log_transform_vars <- c(
  "host_listings_count",
  "price",
  "number_of_reviews"
)

# Transform and Standardise Numeric Modelling Variables:
airbnb_numeric_scaled <- airbnb_reduced |>
  select(all_of(numeric_model_vars)) |>
  mutate(
    across(
      all_of(log_transform_vars),
      ~log1p(.x),
      .names = "log_{.col}"
    )
  ) |>
  select(-all_of(log_transform_vars)) |>
  mutate(
    across(
      everything(),
      ~as.numeric(scale(.x)),
      .names = "scaled_{.col}"
    )
  ) |>
  select(starts_with("scaled_"))

# Rename Variables for Cleaner Modelling Output:
names(airbnb_numeric_scaled) <- names(airbnb_numeric_scaled) |>
  str_remove("^scaled_") |>
  str_remove("^log_")

# Consolidate Categorical Variables Before Dummy Encoding:
n_major_neighbourhoods <- 10
n_major_property_types <- 10

major_neighbourhoods <- airbnb_reduced |>
  count(neighbourhood_cleansed, sort = TRUE) |>
  slice_head(n = n_major_neighbourhoods) |>
  pull(neighbourhood_cleansed)

major_property_types <- airbnb_reduced |>
  count(property_type_proper_trim, sort = TRUE) |>
  slice_head(n = n_major_property_types) |>
  pull(property_type_proper_trim)

airbnb_categorical_consolidated <- airbnb_reduced |>
  mutate(
    neighbourhood_grouped = if_else(
      neighbourhood_cleansed %in% major_neighbourhoods,
      neighbourhood_cleansed,
      "Other"
    ),
    property_type_grouped = if_else(
      property_type_proper_trim %in% major_property_types,
      property_type_proper_trim,
      "Other"
    )
  ) |>
  select(
    neighbourhood_grouped,
    property_type_grouped
  )

# Dummy Encode the Consolidated Categorical Variables:
airbnb_categorical_dummies <- model.matrix(
  ~neighbourhood_grouped + property_type_grouped - 1,
  data = airbnb_categorical_consolidated
) |>
  as.data.frame()

# Clean Dummy Variable Names:
names(airbnb_categorical_dummies) <- names(airbnb_categorical_dummies) |>
  str_replace_all("neighbourhood_grouped", "neighbourhood_") |>
  str_replace_all("property_type_grouped", "property_type_") |>
  str_replace_all("[^A-Za-z0-9_]", "_") |>
  str_replace_all("_+", "_") |>
  str_remove("_$") |>
  str_to_lower()

# Combine Standardised Numeric Variables and Dummy-Encoded Categorical Variables:
airbnb_model_data <- bind_cols(
  airbnb_numeric_scaled,
  airbnb_categorical_dummies
)

# Ensure the Modelling Dataset is Numeric and Complete:
airbnb_model_data <- airbnb_model_data |>
  mutate(across(everything(), as.numeric)) |>
  drop_na()
```

```{r modelling-data-overview}
# Create Modelling Dataset Overview:
model_data_overview <- tibble(
  Dataset = c("Raw Dataset", "Final Modelling Dataset"),
  Observations = c(nrow(airbnb), nrow(airbnb_model_data)),
  Variables = c(ncol(airbnb), ncol(airbnb_model_data))
)

# Output Model Data Overview Table:
model_data_overview_latex <- model_data_overview |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Raw Dataset vs Final Modelling Dataset Used for Clustering",
    align = c("l", "r", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(model_data_overview_latex)
```

## Hierarchical Clustering Sample

Agglomerative hierarchical clustering requires a full pairwise distance matrix. Since the full dataset contains more than sixteen thousand listings, a reproducible sample is used for hierarchical clustering to keep the dendrogram and validation calculations computationally feasible. The sample is used only for the hierarchical clustering analysis.

```{r hc-sample}
# Specify Reproducible Sample Size for Hierarchical Clustering:
hc_sample_size <- min(2500, nrow(airbnb_model_data))

# Draw a Reproducible Sample:
hc_sample_index <- sample(
  seq_len(nrow(airbnb_model_data)),
  size = hc_sample_size
)

# Create HC Modelling and Profiling Datasets:
hc_data <- airbnb_model_data |>
  slice(hc_sample_index)

hc_profile_data <- airbnb_profile_data |>
  slice(hc_sample_index)

# Create Sample Overview Table:
hc_sample_overview <- tibble(
  Dataset = "Hierarchical clustering sample",
  Observations = nrow(hc_data),
  Variables = ncol(hc_data)
)

# Output Sample Overview Table:
hc_sample_overview_latex <- hc_sample_overview |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Dataset Used for Hierarchical Clustering",
    align = c("l", "r", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(hc_sample_overview_latex)
```

## Dissimilarity Matrix

The first step in hierarchical clustering is to define a measure of dissimilarity between observations. Since the final modelling data are numeric and standardised, Euclidean distance is used as the primary dissimilarity measure:

\[ d\_{ii'}=\sqrt{\sum_{j=1}^{p}(x_{ij}-x_{i'j})^2}. \]

```{r hc-distance-matrix}
# Compute Euclidean Distance Matrix:
hc_dist_euclidean <- dist(
  hc_data,
  method = "euclidean"
)
```

## Linkage Method Comparison

Different agglomerative methods define the distance between clusters in different ways. The methods compared here are single linkage, complete linkage, average linkage, centroid linkage and Ward's method. The cophenetic correlation is used to assess how faithfully each dendrogram preserves the original pairwise distances.

```{r hc-linkage-comparison}
# Define Linkage Methods to Compare:
hc_linkage_methods <- c(
  "single",
  "complete",
  "average",
  "centroid",
  "ward.D2"
)

# Fit Hierarchical Clustering Models:
hc_models <- purrr::map(
  hc_linkage_methods,
  \(method_name) {
    hclust(
      hc_dist_euclidean,
      method = method_name
    )
  }
)

# Name the List of Models:
names(hc_models) <- hc_linkage_methods

# Compute Cophenetic Correlations:
hc_linkage_comparison <- tibble(
  Linkage_Method = str_to_title(hc_linkage_methods),
  Cophenetic_Correlation = purrr::map_dbl(
    hc_models,
    \(hc_model) {
      round(
        cor(
          hc_dist_euclidean,
          cophenetic(hc_model)
        ),
        3
      )
    }
  )
) |>
  arrange(desc(Cophenetic_Correlation))

# Output Linkage Comparison Table:
hc_linkage_comparison_latex <- hc_linkage_comparison |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Comparison of Hierarchical Clustering Linkage Methods",
    align = c("l", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(hc_linkage_comparison_latex)
```

```{r hc-dendrogram-comparison, fig.cap="Dendrograms for selected hierarchical linkage methods."}
# Plot Dendrograms for Main Linkage Methods:
par(mfrow = c(2, 2))

plot(
  hc_models[["single"]],
  labels = FALSE,
  main = "Single Linkage",
  xlab = "Listings",
  ylab = "Height"
)

plot(
  hc_models[["complete"]],
  labels = FALSE,
  main = "Complete Linkage",
  xlab = "Listings",
  ylab = "Height"
)

plot(
  hc_models[["average"]],
  labels = FALSE,
  main = "Average Linkage",
  xlab = "Listings",
  ylab = "Height"
)

plot(
  hc_models[["ward.D2"]],
  labels = FALSE,
  main = "Ward.D2 Linkage",
  xlab = "Listings",
  ylab = "Height"
)

par(mfrow = c(1, 1))
```

Although the cophenetic correlation measures distance preservation, the final method is selected with the substantive goal of obtaining compact and interpretable clusters.

```{r hc-final-method}
# Select Final Linkage Method:
hc_final_method <- "ward.D2"

# Extract Final Hierarchical Clustering Model:
hc_final <- hc_models[[hc_final_method]]
```

## Choosing the Number of Clusters

The dendrogram does not automatically determine a single correct number of clusters. Candidate values of (K) are therefore compared using within-cluster variation and average silhouette width. The final cut is based primarily on silhouette width, supported by the dendrogram and interpretability of the cluster profiles.

```{r hc-k-evaluation}
# Candidate Cluster Numbers:
k_values <- 2:8

# Function for Calculating Total Within-Cluster Variation:
within_variation_manual <- function(X, cluster) {
  clusters <- sort(unique(cluster))
  total_within <- 0
  
  for (k in clusters) {
    Xk <- X[cluster == k, , drop = FALSE]
    centroid <- colMeans(Xk)
    squared_distances <- rowSums((sweep(Xk, 2, centroid, "-"))^2)
    total_within <- total_within + sum(squared_distances)
  }
  
  total_within
}

# Evaluate Candidate K Values:
hc_k_evaluation <- map_dfr(
  k_values,
  \(k) {
    cluster_k <- cutree(hc_final, k = k)
    sil_k <- silhouette(cluster_k, hc_dist_euclidean)
    
    tibble(
      K = k,
      Total_Within_Variation = round(
        within_variation_manual(as.matrix(hc_data), cluster_k),
        2
      ),
      Average_Silhouette_Width = round(
        mean(sil_k[, "sil_width"]),
        3
      )
    )
  }
)

# Select K:
hc_best_k = 3

# Output K Evaluation Table:
hc_k_evaluation_latex <- hc_k_evaluation |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Evaluation of Candidate Hierarchical Clustering Solutions",
    align = c("r", "r", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(hc_k_evaluation_latex)
```

```{r hc-k-evaluation-plots, fig.cap="Within-cluster variation and average silhouette width for candidate hierarchical clustering solutions."}
# Set Plot Text Sizes:
plot_title_size <- 14
axis_title_size <- 12
axis_text_size <- 10

# Plot Within-Cluster Variation:
hc_wv_plot <- hc_k_evaluation |>
  ggplot(aes(x = K, y = Total_Within_Variation)) +
  geom_line() +
  geom_point(size = 2) +
  geom_vline(xintercept = hc_best_k, linetype = "dashed") +
  scale_x_continuous(breaks = k_values) +
  labs(
    x = "Number of Clusters (K)",
    y = "Total Within-Cluster Variation"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.title = element_text(size = axis_title_size),
    axis.text = element_text(size = axis_text_size)
  )

# Plot Average Silhouette Width:
hc_sil_plot <- hc_k_evaluation |>
  ggplot(aes(x = K, y = Average_Silhouette_Width)) +
  geom_line() +
  geom_point(size = 2) +
  geom_vline(xintercept = hc_best_k, linetype = "dashed") +
  scale_x_continuous(breaks = k_values) +
  labs(
    x = "Number of Clusters (K)",
    y = "Average Silhouette Width"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.title = element_text(size = axis_title_size),
    axis.text = element_text(size = axis_text_size)
  )

hc_sil_plot
hc_wv_plot
```

## Final Hierarchical Clustering Solution

The final dendrogram is cut at the selected value of (K), and the resulting cluster assignments are used for cluster profiling.

```{r hc-final-clusters}
# Cut the Final Dendrogram:
hc_cluster <- cutree(
  hc_final,
  k = hc_best_k
)

# Add Cluster Assignments to Modelling and Profiling Data:
hc_clustered_data <- hc_data |>
  mutate(
    HC_Cluster = factor(hc_cluster)
  )

hc_profile_clustered <- hc_profile_data |>
  mutate(
    HC_Cluster = factor(hc_cluster)
  )

# Cluster Size Table:
hc_cluster_sizes <- hc_profile_clustered |>
  count(HC_Cluster, name = "N_Listings") |>
  mutate(
    Percentage = percent(N_Listings / sum(N_Listings), accuracy = 0.1)
  )

# Output Cluster Size Table:
hc_cluster_sizes_latex <- hc_cluster_sizes |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Hierarchical Clustering Cluster Sizes",
    align = c("l", "r", "r"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(hc_cluster_sizes_latex)
```

```{r hc-final-dendrogram, fig.cap="Final Ward hierarchical clustering dendrogram with selected cluster cut."}
# Plot Final Dendrogram with Cluster Cut:
plot(
  hc_final,
  labels = FALSE,
  hang = -1,
  main = NULL,
  xlab = "Airbnb Listings",
  ylab = "Height"
)

rect.hclust(
  hc_final,
  k = hc_best_k,
  border = seq_len(hc_best_k) + 1
)
```

```{r hc-final-silhouette, fig.cap="Silhouette plot for the final hierarchical clustering solution."}
# Compute and Plot Final Silhouette Values:
hc_final_silhouette <- silhouette(
  hc_cluster,
  hc_dist_euclidean
)

plot(
  hc_final_silhouette,
  col = seq_len(hc_best_k) + 1,
  border = NA,
  main = "Silhouette Plot for Final HC Solution"
)
```

## Cluster Profiling

The cluster analysis itself was performed on transformed, standardised and dummy-encoded variables. However, cluster interpretation is clearer using the retained variables in their original units.

```{r hc-numeric-cluster-profile}
# Numeric Cluster Profile in Original Units:
hc_numeric_profile <- hc_profile_clustered |>
  group_by(HC_Cluster) |>
  summarise(
    Listings = n(),
    Mean_Price = round(mean(price, na.rm = TRUE), 0),
    Median_Price = round(median(price, na.rm = TRUE), 0),
    Mean_Accommodates = round(mean(accommodates, na.rm = TRUE), 2),
    Mean_Availability_365 = round(mean(availability_365, na.rm = TRUE), 1),
    Mean_Number_of_Reviews = round(mean(number_of_reviews, na.rm = TRUE), 1),
    Mean_Review_Rating = round(mean(review_scores_rating, na.rm = TRUE), 2),
    Mean_Host_Listings_Count = round(mean(host_listings_count, na.rm = TRUE), 1),
    .groups = "drop"
  )

# Output Numeric Cluster Profile:
hc_numeric_profile_latex <- hc_numeric_profile |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Numeric Profile of Hierarchical Clusters in Original Units",
    align = "r",
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(hc_numeric_profile_latex)
```

```{r hc-categorical-cluster-profile}
# Helper Function for Most Common Category:
most_common <- function(x) {
  names(sort(table(x), decreasing = TRUE))[1]
}

# Categorical Cluster Profile:
hc_categorical_profile <- hc_profile_clustered |>
  group_by(HC_Cluster) |>
  summarise(
    Most_Common_Neighbourhood = most_common(neighbourhood_cleansed),
    Most_Common_Property_Type = most_common(property_type_proper_trim),
    .groups = "drop"
  )

# Output Categorical Cluster Profile:
hc_categorical_profile_latex <- hc_categorical_profile |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Dominant Categorical Characteristics of Hierarchical Clusters",
    align = c("l", "l", "l"),
    escape = TRUE
  ) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down")
  )

cat(hc_categorical_profile_latex)
```

```{r hc-standardised-profile-plot, fig.cap="Mean standardised modelling-variable profiles by hierarchical cluster."}
# Create Mean Standardised Profile by Cluster:
hc_standardised_profile <- hc_clustered_data |>
  group_by(HC_Cluster) |>
  summarise(
    across(
      where(is.numeric),
      ~mean(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = -HC_Cluster,
    names_to = "Variable",
    values_to = "Mean_Standardised_Value"
  )

# Plot Standardised Cluster Profiles:
hc_standardised_profile |>
  ggplot(aes(
    x = Variable,
    y = Mean_Standardised_Value,
    group = HC_Cluster,
    colour = HC_Cluster
  )) +
  geom_line() +
  geom_point(size = 1.6) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mean Standardised Value",
    colour = "HC Cluster"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 9),
    axis.title = element_text(size = 11),
    legend.position = "bottom"
  )
```

## PCA Visualisation of the Hierarchical Clusters

PCA is used here only as a two-dimensional visualisation device. The clustering itself was performed on the full model-ready feature space.

```{r hc-pca-visualisation, fig.cap="Two-dimensional PCA visualisation of the final hierarchical clustering solution."}
# Compute PCA Scores for Visualisation:
hc_pca <- prcomp(
  hc_data,
  center = FALSE,
  scale. = FALSE
)

hc_pca_plot_data <- as_tibble(hc_pca$x[, 1:2]) |>
  mutate(
    HC_Cluster = factor(hc_cluster)
  )

# Plot PCA Projection by Cluster:
hc_pca_plot_data |>
  ggplot(aes(x = PC1, y = PC2, colour = HC_Cluster)) +
  geom_point(alpha = 0.65, size = 1.2) +
  labs(
    x = "Principal Component 1",
    y = "Principal Component 2",
    colour = "HC Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```

```{r hc-cluster-spatial-plot, fig.cap="Spatial distribution of hierarchical clusters by latitude and longitude."}
# Plot Hierarchical Clusters by Latitude and Longitude:

# Create Spatial Plot Dataset for the HC Sample:
hc_spatial_plot_data <- airbnb_reduced |>
  slice(hc_sample_index) |>
  mutate(
    HC_Cluster = factor(hc_cluster)
  ) |>
  select(
    longitude,
    latitude,
    HC_Cluster
  )

# Set Plot Text Sizes:
plot_title_size <- 16
axis_title_size <- 13
axis_text_size <- 10
legend_title_size <- 11
legend_text_size <- 10

# Plot Cluster Locations:
hc_spatial_plot_data |>
  ggplot(aes(x = longitude, y = latitude, colour = HC_Cluster)) +
  geom_point(alpha = 0.6, size = 1.2) +
  coord_fixed() +
  labs(
    x = "Longitude",
    y = "Latitude",
    colour = "HC Cluster",
    title = "Spatial Distribution of Hierarchical Clusters"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = plot_title_size, face = "bold"),
    axis.title = element_text(size = axis_title_size),
    axis.text = element_text(size = axis_text_size),
    legend.title = element_text(size = legend_title_size, face = "bold"),
    legend.text = element_text(size = legend_text_size)
  )
```


## Save Hierarchical Clustering Outputs

```{r save-hc-results}
# Save Objects for Later Non-Hierarchical and SOM Comparison:
hc_results <- list(
  sample_index = hc_sample_index,
  model_data = hc_data,
  profile_data = hc_profile_clustered,
  distance_matrix = hc_dist_euclidean,
  linkage_models = hc_models,
  linkage_comparison = hc_linkage_comparison,
  final_method = hc_final_method,
  final_model = hc_final,
  k_evaluation = hc_k_evaluation,
  best_k = hc_best_k,
  clusters = hc_cluster,
  cluster_sizes = hc_cluster_sizes,
  numeric_profile = hc_numeric_profile,
  categorical_profile = hc_categorical_profile,
  silhouette = hc_final_silhouette
)

saveRDS(
  hc_results,
  file = "hc_results_airbnb.rds"
)
```
